---
title: "GATA2_Survey_Data_Exploration"
author: "Ashwin Koppayi"
date: "2025-05-12"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Path Windows
knitr::opts_knit$set(root.dir="/GATA2_Survey/REDCAP_DataExport")

```


```{r}
path_plots<-"/Plots/" 
#Read GATA2 Survey dataframe
gata2_survey_data<-read.csv("GATA2SurveyData_DATA_2025-03-24_1500.csv")

library(ggplot2)
library(readxl)
library(gplots)
library(pheatmap)
library(dplyr)
library(RColorBrewer)
library(pheatmap)
library(reshape2)
library(proxy)
library("klaR")
library(gridExtra)
library(tidyr)
library(forcats)
library(arrow)
library(epitools)
library(stringr)


```



```{r}
#Combining Similar Phenotypes 
gata2_survey_data$lymphatic_vascular_thrombosis_combined <- ifelse(
  gata2_survey_data$lymphatic_vascular_splenic_vein_thrombosis == "Y" |
    gata2_survey_data$lymphatic_vascular_venous_thrombosis == "Y" |
    gata2_survey_data$lymphatic_vascular_thrombotic_venous_events == "Y",
  "Y", "N"
)

#Combining Similar Phenotypes 
gata2_survey_data$lymphatic_vascular_lymphedema_combined <- ifelse(
  gata2_survey_data$head_neck_eyes_lymphedema_limb_or_genital == "Y" |
    gata2_survey_data$lymphatic_vascular_genital_lymphedema == "Y" |
    gata2_survey_data$lymphatic_vascular_primary_uni_bilateral_lymphedema_lower_limb == "Y",
  "Y", "N"
)


#Combining Similar Phenotypes 
gata2_survey_data$viral_infections_ebv_infection_combined <- ifelse(
  gata2_survey_data$viral_infections_ebv_virus == "Y" |
    gata2_survey_data$viral_infections_chronic_active_ebv_disease == "Y" |
    gata2_survey_data$viral_infections_ebv_infection == "Y",
  "Y", "N"
)



#Combining Similar Phenotypes 
gata2_survey_data$viral_infections_hpv_severe_disseminated_genital_extragenital_warts_combined <- ifelse(
  gata2_survey_data$viral_infections_hpv_intraepithelial_bowenoid_papulosis == "Y" |
    gata2_survey_data$viral_infections_hpv_severe_disseminated_genital_extragenital_warts == "Y",
  "Y", "N"
)


#Combining Similar Phenotypes 
gata2_survey_data$cancer_other_cancers <- ifelse(
  gata2_survey_data$cancer_all == "Y" |
    gata2_survey_data$cancer_other_cancers == "Y",
  "Y", "N"
)


#Combining Similar Phenotypes 
gata2_survey_data$head_neck_eyes_sensorineural_hearing_loss <- ifelse(
  gata2_survey_data$head_neck_eyes_congenital_sensorineural_deafness == "Y" |
    gata2_survey_data$head_neck_eyes_treatment_associated_hearing_loss == "Y",
  "Y", "N"
)


#Convert 7.4.1 (Child age 10) preterm labor from "Y" to "NE"
gata2_survey_data[218,153] <- "NE"

#Blank Cells
gata2_survey_data[311,63] <- "NE"
gata2_survey_data[312,63] <- "NE"
gata2_survey_data[169,108] <- "NE"


#Same HGVS and Variant position so changing a variant
#NM_032638.5(GATA2):c.202_203del p.(Val70LeufsTer114) to 
#NM_032638.5(GATA2):c.207_208del p.(Val70LeufsTer114)
gata2_survey_data[286,11]<-"NM_032638.5(GATA2):c.207_208del p.(Val70LeufsTer114)"


```


```{r}
library(readxl)
select_pheno_full<-read_excel("GATA2_Phenotypes_Final_v8.xlsx")
select_pheno<-select_pheno_full[select_pheno_full$Keep_WithZero=="Yes",]


phenotype_gata2<-gata2_survey_data[,38:178]
phenotype_gata2<-phenotype_gata2[,colnames(phenotype_gata2)%in% select_pheno$Phenotype]
rownames(phenotype_gata2)<-gata2_survey_data$prevalence_patient_deidentifier_lab_family_patient
phenotype_gata2$Age<-gata2_survey_data$prevalence_age
phenotype_gata2_adult_binary<-phenotype_gata2[phenotype_gata2$Age>=18,]

phenotype_gata2_adult_binary$Y_count <- rowSums(phenotype_gata2_adult_binary == "Y")
sum(phenotype_gata2_adult_binary$Y_count != 0)

# Count number of "Y" per column
y_counts <- colSums(phenotype_gata2_adult_binary == "Y")

# Turn it into a table (data.frame) for readability
y_counts_table <- data.frame(
  Phenotype = names(y_counts),
  Y_Count   = as.integer(y_counts)
)


# Count number of columns with at least 1 "Y"
sum(y_counts_table$Y_Count != 0)

non_zero<-y_counts_table[y_counts_table$Y_Count!=0,]


select_pheno_107<-select_pheno[select_pheno$Present_in_GATA2=="Yes",]

table(select_pheno_107$Keep_Adults)

```


```{r}
library(dplyr)

# ---- 0.  Handy vectors -------------------------------------------------------
# Identify the phenotype columns (all the columns that start with "P", or list them explicitly)
pheno_cols <- select_pheno$Phenotype   # adjust if needed

# ---- 1.  Flag rows with at least one Y ---------------------------------------
df <- phenotype_gata2 %>%
  mutate(has_any_pheno = if_any(all_of(pheno_cols), ~ .x == "Y"))

# ---- 2.  Overall counts ------------------------------------------------------
total_cases       <- nrow(df)                  # should be 339
total_asymptotic  <- sum(!df$has_any_pheno)    # rows with zero Y’s
total_symptomatic <- total_cases - total_asymptotic

# ---- 3.  Counts by age stratum ----------------------------------------------
age_counts <- df %>%
  mutate(age_group = if_else( Age < 18, "<18", "≥18")) %>%
  group_by(age_group) %>%
  summarise(
    n_participants = n(),
    n_asymptomatic = sum(!has_any_pheno),
    n_symptomatic  = n_participants - n_asymptomatic,
    .groups = "drop"
  )

# ---- 4.  Display -------------------------------------
total_asymptotic
age_counts

```

#Create Files for Combination Analysis
```{r}
phenotype_adult <- phenotype_gata2[phenotype_gata2$Age >= 18, ]

# Initialize an empty list to store participant rownames for each phenotype
phenotype_eid_list <- list()

select_pheno<-select_pheno_full[select_pheno_full$Present_in_GATA2=="Yes" & select_pheno_full$Present_in_UKB=="Yes"&select_pheno_full$Keep_Adults=="Yes",]

  # Loop through each row of select_pheno
for (i in 1:nrow(select_pheno)) {
  phenotype_index <- select_pheno$Phenotype_Index[i]
  col_name <- select_pheno$Phenotype[i]  # Assuming this column contains the actual column name in phenotype_adult
  
  # Check if the column exists in phenotype_adult
  if (col_name %in% colnames(phenotype_adult)) {
    # Get rownames where the value is "Y"
    matching_rows <- rownames(phenotype_adult)[phenotype_adult[[col_name]] == "Y"]
  } else {
    matching_rows <- character(0)  # If column does not exist, leave it empty
  }
  
  # Add to list
  phenotype_eid_list[[phenotype_index]] <- matching_rows
}

# Now convert this list to a dataframe (will need to fill shorter columns with NA)
max_length <- max(sapply(phenotype_eid_list, length))

phenotype_eid_df <- as.data.frame(
  lapply(phenotype_eid_list, function(x) {
    length(x) <- max_length  # Extend with NAs to make equal length
    return(x)
  }),
  stringsAsFactors = FALSE
)

write.csv(phenotype_eid_df,"phenotype_eid_df_Adults_v8.csv",row.names=FALSE)

```


#Create Files for Combination Analysis for Pediatrics
```{r}
phenotype_ped <- phenotype_gata2[phenotype_gata2$Age < 18, ]

# Initialize an empty list to store participant rownames for each phenotype
phenotype_eid_list <- list()

select_pheno<-select_pheno_full[select_pheno_full$Present_in_GATA2=="Yes" & select_pheno_full$Present_in_UKB=="Yes"&select_pheno_full$Keep_Adults=="Yes",]

  # Loop through each row of select_pheno
for (i in 1:nrow(select_pheno)) {
  phenotype_index <- select_pheno$Phenotype_Index[i]
  col_name <- select_pheno$Phenotype[i]  # Assuming this column contains the actual column name in phenotype_adult
  
  # Check if the column exists in phenotype_adult
  if (col_name %in% colnames(phenotype_ped)) {
    # Get rownames where the value is "Y"
    matching_rows <- rownames(phenotype_ped)[phenotype_ped[[col_name]] == "Y"]
  } else {
    matching_rows <- character(0)  # If column does not exist, leave it empty
  }
  
  # Add to list
  phenotype_eid_list[[phenotype_index]] <- matching_rows
}

# Now convert this list to a dataframe (will need to fill shorter columns with NA)
max_length <- max(sapply(phenotype_eid_list, length))

phenotype_eid_df_ped <- as.data.frame(
  lapply(phenotype_eid_list, function(x) {
    length(x) <- max_length  # Extend with NAs to make equal length
    return(x)
  }),
  stringsAsFactors = FALSE
)

write.csv(phenotype_eid_df_ped,"phenotype_eid_df_ped_v8.csv",row.names=FALSE)

```

#Find Coverage of 73 Phenotypes in GATA2
```{r}
# Suppose your dataframe is df
unique_ids <- unique(unlist(phenotype_eid_df))       # Flatten everything and get unique IDs
unique_ids <- unique_ids[!is.na(unique_ids)]   # Remove NAs

# Coverage
(length(unique_ids)/nrow(phenotype_adult))*100

```

```{r}
summary_df <- df %>%
  mutate(
    has_any_pheno = if_any(all_of(pheno_cols), ~ .x == "Y"),
    age_group     = if_else(Age < 18, "<18", "≥18"),
    status        = if_else(has_any_pheno, "Symptomatic", "Asymptomatic")
  ) %>%
  count(age_group, status, name = "n_participants")


# ------------------------------------------------------------------
# 1.  Plot: stacked (or dodge) bar chart ---------------------------
# ------------------------------------------------------------------
ggplot(summary_df,
       aes(x = age_group, y = n_participants, fill = status)) +
  geom_col(position = "stack", width = 0.6) +          # use "dodge" for side‑by‑side
  geom_text(aes(label = n_participants),
            position = position_stack(vjust = 0.5),   # centered on each segment
            size = 4, colour = "black") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title       = "GATA2 cases(n=339) by Age Group",
       x           = "Age group",
       y           = "Number of participants",
       fill        = NULL) +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")

ggsave(paste0(path_plots,"Symptomatic_Asymptomatic_GATA2_By_Age.pdf"))

```

#Single Phenotype Counts comparison
```{r}
phenotype_ukb<-read_excel("Phenotype_pid_final_022526_v8.xlsx")
single_pheno<-data.frame(Phenotype_Index=colnames(phenotype_eid_df))
single_pheno<-merge(single_pheno,select_pheno[,c("Phenotype_Index","Phenotype_Description_Expanded","New_Category")],by="Phenotype_Index")

single_pheno$Counts_GATA2<-rep(NA,nrow(single_pheno))
single_pheno$Counts_UKB<-rep(NA,nrow(single_pheno))
for(i in 1:nrow(single_pheno)){
  print(i)
  single_pheno$Counts_GATA2[i]<-sum(!is.na(phenotype_eid_df[,single_pheno$Phenotype_Index[i]]))
    single_pheno$Counts_UKB[i]<-length(which(phenotype_ukb[,single_pheno$Phenotype_Index[i]]!=0))
}

write.csv(single_pheno,"SinglePheno_GATA2_v8.csv",row.names=FALSE)

```



#Genotype Summary
```{r}


variant_type<-read_excel("GATA2_Variants_Final.xlsx")



#Same HGVS and Variant position so changing a variant
#NM_032638.5(GATA2):c.202_203del p.(Val70LeufsTer114) to 
#NM_032638.5(GATA2):c.207_208del p.(Val70LeufsTer114)
gata2_survey_data[286,11]<-"NM_032638.5(GATA2):c.207_208del p.(Val70LeufsTer114)"


variant_counts <- as.data.frame(table(variant_type$VariantType))

# Rename the columns for clarity
colnames(variant_counts) <- c("Variant_Type", "Count")



# Calculate the percentage
variant_counts$Percentage <- round(variant_counts$Count / sum(variant_counts$Count) * 100, 1)

# Reorder the Variant_Type factor by decreasing percentage
variant_counts <- variant_counts[order(-variant_counts$Percentage), ]
variant_counts$Variant_Type <- factor(variant_counts$Variant_Type, levels = variant_counts$Variant_Type)

variant_counts <- variant_counts %>%
  mutate(pos = cumsum(Count) - Count / 2)
variant_counts$Variant_Type<-gsub("_"," ",variant_counts$Variant_Type)

library(dplyr)

variant_counts <- variant_counts %>%
  arrange(desc(Percentage), desc(Count)) %>%              # tie-break by Count if needed
  mutate(Variant_Type = factor(Variant_Type, levels = Variant_Type))

pdf(paste0(path_plots, "VariantType_Piechart1.pdf"))
ggplot(variant_counts, aes(x = "", y = Count, fill = Variant_Type)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y") +
  labs(title = "Distribution of Variant Types", fill = "Variant Type") +
  geom_text(aes(label = paste0(Percentage, "%")),
            position = position_stack(vjust = 0.5)) +
  theme_void()
dev.off()


gata2_survey_data <- gata2_survey_data %>%
  dplyr::left_join(
    variant_type %>% dplyr::select(Var1, VariantType),
    by = c("prevalence_variant_name" = "Var1")
  )






```


#Frequency of Phenotype (Only the One's with >0 frequency,Expanded)
```{r}
# ─────────────────────────────────────────────────────────────
# 0.  Libraries  (install with install.packages() if needed)
# ─────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(ggplot2)

# ─────────────────────────────────────────────────────────────
# 1.  Identify phenotype columns
#     ⇨ tweak the grep() pattern if your column names differ
# ─────────────────────────────────────────────────────────────
mapping_df <- select_pheno %>%                # or read.csv("…")
  dplyr::select(Phenotype, Phenotype_Description_Expanded,New_Category)
colnames(mapping_df)[1]<-"Phenotype"
pheno_cols <- mapping_df$Phenotype

# ─────────────────────────────────────────────────────────────
# 2.  Build a tidy table of counts per phenotype × age group
# ─────────────────────────────────────────────────────────────
counts_df <- phenotype_gata2 %>%
  mutate(age_group = if_else(Age < 18, "<18", "≥18")) %>%
  pivot_longer(cols = all_of(pheno_cols),
               names_to  = "Phenotype",
               values_to = "value") %>%
  filter(value == "Y") %>%               # keep only present phenotypes
  count(age_group, Phenotype, name = "n")%>%
  left_join(mapping_df, by = "Phenotype")  

# ─────────────────────────────────────────────────────────────
# 3.  Define a tiny helper that returns a ggplot for one stratum
# ─────────────────────────────────────────────────────────────
make_plot <- function(.data, group_label, top_n = 200) {
  .data %>%
    filter(age_group == group_label) %>%
    slice_max(order_by = n, n = top_n, with_ties = FALSE) %>%
    ggplot(aes(x = reorder(Phenotype_Description_Expanded, n), y = n, fill = New_Category)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = n), hjust = -0.15, size = 4.0) +
      coord_flip(clip = "off") +
    scale_fill_brewer(palette = "Set3") +
      scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
      labs(title = paste0("Frequency of Phenotypes in ", group_label, " years"),
           x     = NULL,
           y     = "Number of participants",
           fill  = "Category") +
      theme_classic(base_size = 13)+
    theme(axis.text.y = element_text(size = 11,face="bold") )
}

# ─────────────────────────────────────────────────────────────
# 4.  Create the two plots
# ─────────────────────────────────────────────────────────────
plot_under18 <- make_plot(counts_df, "<18")
plot_adult   <- make_plot(counts_df, "≥18")

# ─────────────────────────────────────────────────────────────
# 5.  Save each to its own PDF
# ─────────────────────────────────────────────────────────────
ggsave(paste0(path_plots,"phenotypes_under18_nonzero_only_expanded.pdf"),
       plot   = plot_under18,
       width  = 12, height = 15, units = "in")

ggsave(paste0(path_plots,"phenotypes_adults_nonzero_only_expanded.pdf"),
       plot   = plot_adult,
       width  = 12, height = 15, units = "in")

```

#Frequency of Phenotype (Only the One's with >0 frequency,v2)
```{r}
# ─────────────────────────────────────────────────────────────
# 0.  Libraries  (install with install.packages() if needed)
# ─────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(ggplot2)

# ─────────────────────────────────────────────────────────────
# 1.  Identify phenotype columns
#     ⇨ tweak the grep() pattern if your column names differ
# ─────────────────────────────────────────────────────────────
mapping_df <- select_pheno %>%                # or read.csv("…")
  dplyr::select(Phenotype, Phenotype_Description_v2,New_Category)
colnames(mapping_df)[1]<-"Phenotype"
pheno_cols <- mapping_df$Phenotype

# ─────────────────────────────────────────────────────────────
# 2.  Build a tidy table of counts per phenotype × age group
# ─────────────────────────────────────────────────────────────
counts_df <- phenotype_gata2 %>%
  mutate(age_group = if_else(Age < 18, "<18", "≥18")) %>%
  pivot_longer(cols = all_of(pheno_cols),
               names_to  = "Phenotype",
               values_to = "value") %>%
  filter(value == "Y") %>%               # keep only present phenotypes
  count(age_group, Phenotype, name = "n")%>%
  left_join(mapping_df, by = "Phenotype")  

# ─────────────────────────────────────────────────────────────
# 3.  Define a tiny helper that returns a ggplot for one stratum
# ─────────────────────────────────────────────────────────────
make_plot <- function(.data, group_label, top_n = 200) {
  .data %>%
    filter(age_group == group_label) %>%
    slice_max(order_by = n, n = top_n, with_ties = FALSE) %>%
    ggplot(aes(x = reorder(Phenotype_Description_v2, n), y = n, fill = New_Category)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = n), hjust = -0.15, size = 4.0) +
      coord_flip(clip = "off") +
    scale_fill_brewer(palette = "Set3") +
      scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
      labs(title = paste0("Frequency of Phenotypes in ", group_label, " years"),
           x     = "Observed Phenotypes",
           y     = "Number of participants",
           fill  = "Category") +
      theme_classic(base_size = 13)+
    theme(axis.text.y = element_text(size = 11,face="plain"))
}

# ─────────────────────────────────────────────────────────────
# 4.  Create the two plots
# ─────────────────────────────────────────────────────────────
plot_under18 <- make_plot(counts_df, "<18")
plot_adult   <- make_plot(counts_df, "≥18")

# ─────────────────────────────────────────────────────────────
# 5.  Save each to its own PDF
# ─────────────────────────────────────────────────────────────
ggsave(paste0(path_plots,"phenotypes_under18_nonzero_only_expanded_v2.pdf"),
       plot   = plot_under18,
       width  = 12, height = 15, units = "in")

ggsave(paste0(path_plots,"phenotypes_adults_nonzero_only_expanded_v2.pdf"),
       plot   = plot_adult,
       width  = 12, height = 15, units = "in")

```



#Frequency of 73 Phenotypes used in combination analysis (frequency,v2)
```{r}
# ─────────────────────────────────────────────────────────────
# 0.  Libraries  (install with install.packages() if needed)
# ─────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(ggplot2)

# ─────────────────────────────────────────────────────────────
# 1.  Identify phenotype columns
#     ⇨ tweak the grep() pattern if your column names differ
# ─────────────────────────────────────────────────────────────
mapping_df <- select_pheno[select_pheno$Phenotype_Index%in%select_pheno$Phenotype_Index,] %>%                # or read.csv("…")
  dplyr::select(Phenotype, Phenotype_Description_v2,New_Category)

colnames(mapping_df)[1]<-"Phenotype"
pheno_cols <- mapping_df$Phenotype

# ─────────────────────────────────────────────────────────────
# 2.  Build a tidy table of counts per phenotype × age group
# ─────────────────────────────────────────────────────────────
counts_df <- phenotype_gata2 %>%
  mutate(age_group = if_else(Age < 18, "<18", "≥18")) %>%
  pivot_longer(cols = all_of(pheno_cols),
               names_to  = "Phenotype",
               values_to = "value") %>%
  filter(value == "Y") %>%               # keep only present phenotypes
  count(age_group, Phenotype, name = "n")%>%
  left_join(mapping_df, by = "Phenotype")  

# ─────────────────────────────────────────────────────────────
# 3.  Define a tiny helper that returns a ggplot for one stratum
# ─────────────────────────────────────────────────────────────
make_plot <- function(.data, group_label, top_n = 200) {
  .data %>%
    filter(age_group == group_label) %>%
    slice_max(order_by = n, n = top_n, with_ties = FALSE) %>%
    ggplot(aes(x = reorder(Phenotype_Description_v2, n), y = n, fill = New_Category)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = n), hjust = -0.15, size = 4.0) +
      coord_flip(clip = "off") +
    scale_fill_brewer(palette = "Set3") +
      scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
      labs(
           x     = "Observed phenotypes",
           y     = "Number of participants",
           fill  = "Category") +
      theme_classic(base_size = 13)+
    theme(axis.text.y = element_text(size = 11,face="plain") ) +
  theme(legend.position = c(0.5, 0.2))
}

# ─────────────────────────────────────────────────────────────
# 4.  Create the two plots
# ─────────────────────────────────────────────────────────────
plot_under18 <- make_plot(counts_df, "<18")
plot_adult   <- make_plot(counts_df, "≥18")

# ─────────────────────────────────────────────────────────────
# 5.  Save each to its own PDF
# ─────────────────────────────────────────────────────────────
ggsave(paste0(path_plots,"73phenotypes_under18_nonzero_only_expanded_v2.pdf"),
       plot   = plot_under18,
       width  = 12, height = 15, units = "in")

ggsave(paste0(path_plots,"73phenotypes_adults_nonzero_only_expanded_v2.pdf"),
       plot   = plot_adult,
       width  = 12, height = 15, units = "in")

```




#Including Zero 

```{r}

# ─────────────────────────────────────────────────────────────
# 2.  Build a tidy table of counts per phenotype × age group
# ─────────────────────────────────────────────────────────────
# All possible combinations of age_group × phenotype
all_combos <- expand.grid(
  age_group = c("<18", "≥18"),
  Phenotype = pheno_cols,
  stringsAsFactors = FALSE
)

counts_df <- phenotype_gata2 %>%
  mutate(age_group = if_else(Age < 18, "<18", "≥18")) %>%
  pivot_longer(cols = all_of(pheno_cols),
               names_to  = "Phenotype",
               values_to = "value") %>%
  filter(value == "Y") %>%
  count(age_group, Phenotype, name = "n") %>%
  right_join(all_combos, by = c("age_group", "Phenotype")) %>%   # keep 0 counts
  mutate(n = replace_na(n, 0)) %>%
  left_join(mapping_df, by = "Phenotype")

# ─────────────────────────────────────────────────────────────
# 3.  Define a tiny helper that returns a ggplot for one stratum
# ─────────────────────────────────────────────────────────────
make_plot <- function(.data, group_label, top_n = 200) {
  .data %>%
    filter(age_group == group_label) %>%
    slice_max(order_by = n, n = top_n, with_ties = FALSE) %>%
    ggplot(aes(x = reorder(Phenotype_Description_v2, n), y = n, fill = New_Category)) +
      geom_col(width = 0.7) +
      geom_text(aes(label = n), hjust = -0.15, size = 3.5) +
      coord_flip(clip = "off") +
    scale_fill_brewer(palette = "Set3") +
      scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
      labs(title = paste0("Frequency of Phenotypes in ", group_label, " years"),
           x     = NULL,
           y     = "Number of participants",
           fill  = "Category") +
      theme_classic(base_size = 12)
}

# ─────────────────────────────────────────────────────────────
# 4.  Create the two plots
# ─────────────────────────────────────────────────────────────
plot_under18 <- make_plot(counts_df, "<18")
plot_adult   <- make_plot(counts_df, "≥18")

# ─────────────────────────────────────────────────────────────
# 5.  Save each to its own PDF
# ─────────────────────────────────────────────────────────────
ggsave(paste0(path_plots,"phenotypes_under18_withall.pdf"),
       plot   = plot_under18,
       width  = 12, height = 15, units = "in")

ggsave(paste0(path_plots,"phenotypes_adults_withall.pdf"),
       plot   = plot_adult,
       width  = 12, height = 15, units = "in")

```




#Combined Frequency
```{r}
# ─────────────────────────────────────────────────────────────
# 4-bis.  One combined plot: adults vs <18, ordered by adults
# ─────────────────────────────────────────────────────────────
top_n <- 200          # change or remove this line to keep more/fewer phenotypes

# 4-bis-A.  Get the order based on the adult stratum
adult_order <- counts_df %>%
  filter(age_group == "≥18") %>%                 # keep adults only
  arrange(n) %>%                           # sort by descending count
  { if (!is.null(top_n)) slice_head(., n = top_n) else . } %>%  # optional top-n
  pull(Phenotype_Description_v2)

# 4-bis-B.  Build the plot
plot_combined <- counts_df %>%
  filter(Phenotype_Description_v2 %in% adult_order) %>%            # keep same set in both groups
  mutate(Phenotype_Description = factor(Phenotype_Description_v2,
                                        levels = adult_order)) %>%  # set factor order
  ggplot(aes(x = Phenotype_Description_v2, y = n, fill = age_group)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    #geom_text(aes(label = n),
    #          position = position_dodge(width = 0.8),
     #         hjust = -0.15, size = 3.5) +
    coord_flip(clip = "off") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    scale_fill_manual(values = c("≥18" = "#1f77b4", "<18" = "#ff7f0e"),
                      name   = "Age group") +
    labs(title = "Frequency of phenotypes by age group",
         x     = NULL,
         y     = "Number of participants") +
    theme_classic(base_size = 12)

# ─────────────────────────────────────────────────────────────
# 4-bis-C.  Save the figure
# ─────────────────────────────────────────────────────────────
ggsave(paste0(path_plots,"phenotypes_combined_top200.pdf"),
       plot   = plot_combined,
       width  = 12, height = 15, units = "in")

```


#Original Plot Showing the NE values

#Summary of Frequency of phenotypes
```{r}
#Remove the 
phenotype_gata2_og<-phenotype_gata2
phenotype_gata2<-phenotype_gata2 %>% dplyr::select(-Age)

#Initlize empty dataframe for getting the frequency of phenotypes
frequency<-data.frame(matrix(0,nrow=ncol(phenotype_gata2),ncol=3))
colnames(frequency)<-c("Y","N","NE")


for( i in 1:length(colnames(phenotype_gata2))){
#print(i)
  data<-as.data.frame(table(phenotype_gata2[,i]))
  row.names(frequency)[i]<-colnames(phenotype_gata2[i])
  frequency$Y[i]<-ifelse(identical(data$Freq[data$Var1=="Y"],integer(0)),0,data$Freq[data$Var1=="Y"])
  frequency$N[i]<-ifelse(identical(data$Freq[data$Var1=="N"],integer(0)),0,data$Freq[data$Var1=="N"])
  frequency$NE[i]<-ifelse(identical(data$Freq[data$Var1=="NE"],integer(0)),0,data$Freq[data$Var1=="NE"])
  
}

library(ggplot2)
library(reshape2)
library(tibble)

mapping_df <- select_pheno %>%                # or read.csv("…")
  dplyr::select(Phenotype, Phenotype_Description_v2,New_Category)
colnames(mapping_df)[1]<-"Phenotype"
pheno_cols <- mapping_df$Phenotype


frequency_short<-frequency[,1:3]
frequency_short<-rownames_to_column(frequency_short,var="Phenotype")
colnames(frequency_short)<-c("Phenotype","Y","N","NE")
frequency_long <- melt(frequency_short, id.vars = "Phenotype", 
                  variable.name = "Response", 
                  value.name = "Frequency")
frequency_short$Total<-frequency_short$Y+ frequency_short$N +frequency_short$NE

# Adjust the factor levels
frequency_long$Response <- factor(frequency_long$Response, levels = c("Y","N","NE"))

frequency_long<-merge(frequency_long,mapping_df,by="Phenotype")


```


```{r}
library(dplyr)
library(ggplot2)

## 1.  Get the “Y” counts per phenotype and rank them
order_levels <- frequency_long %>% 
  filter(Response %in% c("Frequency_Y", "Y")) %>%   # adjust if your Y code is just "Y"
  group_by(Phenotype_Description_v2) %>%               # one phenotype per line
  summarise(Y_freq = sum(Frequency), .groups = "drop") %>% 
  arrange(desc(Y_freq)) %>%                         # highest Y first
  pull(Phenotype_Description_v2)                       # keep only the names

## 2.  Apply that order to the factor
frequency_long$Phenotype_Description_v2 <- factor(
  frequency_long$Phenotype_Description_v2,
  levels = order_levels
)


# 100 % color‑blind‑friendly (Okabe‑Ito)
cb_palette <- c(
  Y  = "#0072B2",   # blue
  N  = "#D55E00",   # vermillion
  NE = "#F0E442"    # yellow
)

pdf(paste0(path_plots,"Frequency_phenotype_stacked_v2_final.pdf"),width=15,height=20)
ggplot(frequency_long,
       aes(x = Phenotype_Description_v2, y = Frequency, fill = Response)) +
  geom_col(color = "black",
           position = position_stack(reverse = TRUE, vjust = 1.05)) +
  scale_fill_manual(values = cb_palette,      # ⬅️  new line
                    name   = "Response",
                    labels = c(Frequency_Y = "Y", Frequency_N = "N", Frequency_NE = "NE")) +
  labs(
       x = "Phenotype",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1,size = 11),panel.grid.major = element_blank(),
  panel.grid.minor = element_blank())
dev.off()
```



#Missingness by Site
```{r}
site_missingness<-cbind(gata2_survey_data$prevalence_lab_expert,phenotype_gata2)

colnames(site_missingness)[1]<-"Prevalence_lab_expert"

# Convert Category to a factor to control the order of bars
site_missingness$Prevalence_lab_expert <- factor(site_missingness$Prevalence_lab_expert , levels = rev(unique(site_missingness$Prevalence_lab_expert)))

site_missingness$Prevalence_lab_expert<-gsub("14. NIH ","14. NIH",site_missingness$Prevalence_lab_expert)

summary(as.factor(site_missingness$Prevalence_lab_expert))


# Convert frequency table to dataframe
summary_df <- as.data.frame(table(site_missingness$Prevalence_lab_expert))
summary_df <- summary_df[order(-summary_df$Freq), ]
summary_df$Var1<-as.character(summary_df$Var1)

site_map_file<-read.csv("Site_Map_File.csv")





summary_df<-merge(summary_df,site_map_file,by.x="Var1",by.y="Site_GATA2",all=TRUE)

library(dplyr)

summary_grouped <- summary_df %>%
  group_by(Site) %>%
  summarise(Freq = sum(Freq), .groups = "drop")

write.csv(summary_df,"Site_Map_File_Summary.csv",row.names=FALSE)
# Plot using ggplot2
library(ggplot2)
library(dplyr)
library(forcats)



summary_grouped <- summary_df %>%
  dplyr::mutate(Site = str_trim(Site)) %>%  # Trim leading/trailing spaces
  dplyr::group_by(Site) %>%
  summarise(Freq = sum(Freq), .groups = "drop")
# Ensure Site is ordered by Freq decreasing
summary_grouped <- summary_grouped %>%
  mutate(Site = fct_reorder(Site, Freq, .desc = TRUE))

  pdf(paste0(path_plots,"Number_of_GATA2_Cases_by_Site.pdf"),width=8,height=6)
  ggplot(summary_grouped, aes(x = Site, y = Freq, fill = Site)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = Freq), vjust = -0.5, size = 3) +
    labs(x = "Site", y = "Frequency", title = "Number of GATA2 Deficient Cases at each site") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  dev.off()
```
```{r}
# (optional) make sure keys are comparable
summary_df$Var1 <- trimws(as.character(summary_df$Var1))
site_missingness$Prevalence_lab_expert <- trimws(as.character(site_missingness$Prevalence_lab_expert))

# add Site by matching keys
site_missingness$Site <- summary_df$Site[
  match(site_missingness$Prevalence_lab_expert, summary_df$Var1)
]

```



```{r}
library(dplyr)
library(stringr)

# Clean keys
summary_df <- summary_df %>%
  mutate(Site = str_trim(as.character(Site)),
         Var1 = str_trim(as.character(Var1)))

site_missingness <- site_missingness %>%
  mutate(Prevalence_lab_expert = str_trim(as.character(Prevalence_lab_expert)))

# 1) One row per Site with total n (Freq)
summary_df_group <- summary_df %>%
  group_by(Site) %>%
  summarise(Freq = sum(Freq, na.rm = TRUE), .groups = "drop")

cat("Unique sites:", nrow(summary_df_group), "\n")

# Optional: define label order to match your 3 colors (edit if your labels differ)
# expected_levels <- c("Not Missing", "Missing", "Unknown")

# 2) Plot per-site pies
pdf(paste0(path_plots, "Combined_bar_plot.pdf"),width=7,height=8)

n_sites <- nrow(summary_df_group)
n_cols  <- 3
n_rows  <- ceiling(n_sites / n_cols)
par(mfrow = c(n_rows, n_cols), cex.axis = 1)

for (i in seq_len(n_sites)) {
  site_name <- summary_df_group$Site[i]
  n_value   <- summary_df_group$Freq[i]

  # All Var1 for this Site
  var1_list <- summary_df %>%
    filter(Site == site_name) %>%
    pull(Var1) %>%
    unique()

  # Subset missingness rows for those Var1 values
  missing_by_site <- site_missingness %>%
    filter(Prevalence_lab_expert %in% var1_list) %>%
    dplyr::select(-c(Prevalence_lab_expert,Site))

  # Flatten to a vector of labels; drop NA/blank
  vals <- unlist(missing_by_site, use.names = FALSE)
  vals <- vals[!is.na(vals) & vals != ""]

  if (length(vals) > 0) {
    frequency_site <- table(vals)

    # If you want fixed order & to align colors, uncomment the two lines below and set expected_levels above
    # frequency_site <- table(factor(vals, levels = expected_levels))
    # frequency_site <- frequency_site[frequency_site > 0]

    pie(
      frequency_site,
      labels = paste0(
        names(frequency_site), ": ",
        round(100 * as.numeric(frequency_site) / sum(frequency_site), 2), "%"
      ),
      main = paste0(site_name, "\n", "n = ", n_value),
      col  = c("#D55E00","#F0E442","#0072B2")  # adjust if you change label order
    )
  } else {
    plot.new()
    text(1, 1, paste0(site_name, "\nNo data"))
  }
}

par(mfrow = c(1, 1))
dev.off()

```



```{r}
# --- libraries ---
library(dplyr)
library(stringr)

# --- tweakables ---
PIE_RADIUS   <- 0.80   # pie size (smaller = more margin)
LABEL_RADIUS <- 1.15   # >1 moves labels away from the pie
LABEL_CEX    <- 0.8    # label text size
LABEL_COLORS <- c("#D55E00", "#F0E442", "#0072B2")  # make sure order matches your levels

# --- clean keys ---
summary_df <- summary_df %>%
  mutate(
    Site = str_trim(as.character(Site)),
    Var1 = str_trim(as.character(Var1))
  )

site_missingness <- site_missingness %>%
  mutate(Prevalence_lab_expert = str_trim(as.character(Prevalence_lab_expert)))

# --- one row per Site with total n (Freq) ---
summary_df_group <- summary_df %>%
  group_by(Site) %>%
  summarise(Freq = sum(Freq, na.rm = TRUE), .groups = "drop")

cat("Unique sites:", nrow(summary_df_group), "\n")

# --- output device ---
pdf(paste0(path_plots, "Combined_bar_plot_o2.pdf"), width = 8, height = 10)

n_sites <- nrow(summary_df_group)
n_cols  <- 3
n_rows  <- ceiling(n_sites / n_cols)
par(mfrow = c(n_rows, n_cols), cex.axis = 1)

for (i in seq_len(n_sites)) {
  site_name <- summary_df_group$Site[i]
  n_value   <- summary_df_group$Freq[i]

  # All Var1 for this Site
  var1_list <- summary_df %>%
    filter(Site == site_name) %>%
    pull(Var1) %>%
    unique()

  # Subset missingness rows for those Var1 values
  missing_by_site <- site_missingness %>%
    filter(Prevalence_lab_expert %in% var1_list) %>%
    dplyr::select(-c(Prevalence_lab_expert, Site))

  # Flatten to a vector of labels; drop NA/blank
  vals <- unlist(missing_by_site, use.names = FALSE)
  vals <- vals[!is.na(vals) & vals != ""]

  if (length(vals) > 0) {
    frequency_site <- table(vals)

    # (Optional) enforce a fixed order to match colors:
    # expected_levels <- c("Not Missing", "Missing", "Unknown")
    # frequency_site <- table(factor(vals, levels = expected_levels))
    # frequency_site <- frequency_site[frequency_site > 0]

    # Build label text
    pct <- round(100 * as.numeric(frequency_site) / sum(frequency_site), 2)
    pie_labels <- paste0(names(frequency_site), ": ", pct, "%")

    # Draw pie WITHOUT labels first (we'll place them manually)
    pie(
      frequency_site,
      labels = NA,
      main   = paste0(site_name, "\n", "n = ", n_value),
      col    = LABEL_COLORS,
      radius = PIE_RADIUS
    )

    # Compute mid-angles of slices (in radians)
    mids <- cumsum(frequency_site) - frequency_site / 2
    mids <- 2 * pi * mids / sum(frequency_site)

    # Place labels slightly outside the pie
    x <- LABEL_RADIUS * sin(mids)
    y <- LABEL_RADIUS * cos(mids)
    text(x, y, labels = pie_labels, cex = LABEL_CEX)

  } else {
    plot.new()
    text(0.5, 0.5, paste0(site_name, "\nNo data"))
  }
}

par(mfrow = c(1, 1))
dev.off()

```

```{r}
# --- Libraries ---
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)

# --- Clean ---
summary_df <- summary_df %>%
  mutate(
    Site = str_trim(as.character(Site)),
    Var1 = str_trim(as.character(Var1))
  )

site_missingness <- site_missingness %>%
  mutate(Prevalence_lab_expert = str_trim(as.character(Prevalence_lab_expert)))

# --- Map each Var1 to its Site(s) ---
var_by_site <- summary_df %>%
  distinct(Site, Var1)

# --- Identify phenotype columns (all Y/N/NE columns) ---
response_cols <- setdiff(names(site_missingness), c("Prevalence_lab_expert", "Site"))

# --- Join safely ---
long_df <- site_missingness %>%
  # Keep only phenotype columns and the join key
  dplyr::select(Prevalence_lab_expert, all_of(response_cols)) %>%
  inner_join(var_by_site, by = c("Prevalence_lab_expert" = "Var1")) %>%
  pivot_longer(
    cols = all_of(response_cols),
    names_to = "Phenotype",
    values_to = "Response"
  ) %>%
  filter(Response %in% c("Y", "N", "NE")) %>%
  dplyr::select(Site, Phenotype, Response)
# --- Count and percentage per site ---
site_counts <- long_df %>%
  count(Site, Response, name = "Count") %>%
  group_by(Site) %>%
  mutate(Total = sum(Count),
         Percent = 100 * Count / Total) %>%
  ungroup()

# --- Ensure consistent ordering of responses ---
site_counts <- site_counts %>%
  mutate(Response = factor(Response, levels = c("N", "NE", "Y")))

pdf(paste0(path_plots, "MissingNess_By_Site.pdf"), width = 8, height = 6)
# --- Plot stacked horizontal barplot ---
ggplot(site_counts, aes(x = Percent, y = reorder(Site, Total), fill = Response)) +
  geom_col(width = 0.7) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 10),
                     labels = function(x) paste0(x, "%")) +
  scale_fill_manual(values = c("N" = "#D55E00", "NE" = "#F0E442", "Y" = "#0072B2")) +
  labs(
    x = "Percentage of Response",
    y = "Site",
    fill = "Response",
      ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "top",
    panel.grid.major=element_blank(),
    panel.grid.minor=element_blank()
  )

dev.off()

```





```{r}
# ─────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(pheatmap)
library(RColorBrewer)

# ─────────────────────────────────────────────────────────────
# 1.  Mapping table: code  →  description
# ─────────────────────────────────────────────────────────────

mapping_df <- select_pheno %>% dplyr::select(Phenotype, Phenotype_Description)

# ─────────────────────────────────────────────────────────────
# 2.  Identify phenotype columns present in the main data frame
# ─────────────────────────────────────────────────────────────
pheno_cols <- mapping_df$Phenotype         # e.g. "P1"…"P77"


wide_df <- phenotype_gata2_og %>%
  dplyr::select(Age, all_of(pheno_cols)) %>%
  mutate(across(all_of(pheno_cols), ~ ifelse(.x == "Y", "Y", ifelse(.x == "N", "N", "NE")))) %>%
  mutate(age_group = if_else(Age < 18, "<18", "≥18"))

heat_mat <- wide_df %>%
  dplyr::select(all_of(pheno_cols)) %>%
  as.matrix()

colnames(heat_mat) <- mapping_df$Phenotype_Description[
  match(colnames(heat_mat), mapping_df$Phenotype)
]

heatmap_colors <- c(
  "Y"  = "#0072B2",    # Blue
  "N"  = "#D55E00",    # Vermillion
  "NE" = "#FFFFFF"     # White
)
# Map Y/N/NE to 1/2/3 for plotting
heat_mat_numeric <- matrix(
  dplyr::recode(heat_mat, Y = 1L, N = 2L, NE = 3L),
  nrow = nrow(heat_mat),
  ncol = ncol(heat_mat),
  dimnames = dimnames(heat_mat)
)
# set up color vector
my_colors <- c("#0072B2", "#D55E00", "#FFFFFF")
names(my_colors) <- c("1", "2", "3")

heat_mat_numeric_t <- t(heat_mat_numeric)

pdf(paste0(path_plots,"Heatmap_Combined_normal.pdf"), width = 15, height =11 )
pheatmap(
  heat_mat_numeric,
  color = my_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,

  annotation_colors = list(Age = c("<18" = "#fc8d59", "≥18" = "#99d594")),
 
  show_rownames = FALSE,
  fontsize_col = 8,
  main = "Phenotype presence by age group (n = 339)"
)
dev.off()
```

```{r}
heat_mat_og<-heat_mat
heat_mat_df <- as.data.frame(heat_mat)
heat_mat_df$ParticipantID <- rownames(heat_mat)
# Merge variant type info
heat_mat_df <- heat_mat_df %>%
  left_join(
    gata2_survey_data %>% dplyr::select(prevalence_patient_deidentifier_lab_family_patient, VariantType),
    by = c("ParticipantID" = "prevalence_patient_deidentifier_lab_family_patient")
  )
```



```{r}
library(dplyr)
library(tidyr)

heat_long <- heat_mat_df %>%
  pivot_longer(
    cols = -ParticipantID,        # all phenotype columns
    names_to = "Phenotype",
    values_to = "Status"
  )

heat_long <- heat_long %>%
  left_join(
    gata2_survey_data[, c("prevalence_patient_deidentifier_lab_family_patient", "VariantType")],
    by = c("ParticipantID" = "prevalence_patient_deidentifier_lab_family_patient")
  )


heat_long_unique <- heat_long %>%
  filter(Status == "Y") %>%
  group_by(ParticipantID, Phenotype) %>%
  summarise(
    VariantType = paste(unique(VariantType), collapse = "; "),
    .groups = "drop"
  )

# Now pivot wider safely:
heat_wide <- heat_long_unique %>%
  pivot_wider(
    names_from = Phenotype,
    values_from = VariantType
  )

# Make matrix:
heatmap_matrix <- heat_wide %>%
  column_to_rownames("ParticipantID") %>%
  as.matrix()

heatmap_matrix[is.na(heatmap_matrix)] <- ""




```




#Heatmap Oncoprint

```{r}
library(ComplexHeatmap)
library(grid)

# Define your mutation colors properly
mutation_colors <- c(
  "Missense_Mutation" = "#E64B35",
  "Nonsense_Mutation" = "#4DBBD5",
  "Large_Deletion" = "#7E6148",
  "Inframe_Insertion" = "#B09C85",
  "Frameshift_Deletion" = "#FF7F00",
  "Noncoding_Mutation" = "#6A3D9A",
  "Frameshift_Insertion" = "#1F78B4",
  "Indel" = "#A6CEE3",
  "Synonymous" = "#B2DF8A",
  "Inframe_Deletion" = "#33A02C"
)

# Define background
alter_fun <- list(
  background = function(x, y, w, h) {
    grid.rect(x, y, w, h, gp = gpar(fill = "grey90", col = NA))
  }
)

# Correctly capture color inside loop using local()
for (mutation in names(mutation_colors)) {
  alter_fun[[mutation]] <- local({
    mut <- mutation  # capture current mutation name
    function(x, y, w, h) {
      grid.rect(x, y, w * 0.9, h * 0.9, gp = gpar(fill = mutation_colors[[mut]], col = NA))
    }
  })
}


library(ComplexHeatmap)
library(grid)

# count “any alteration present” per row (adjust if your matrix stores data differently)
mut_counts <- rowSums(heatmap_matrix != "" & !is.na(heatmap_matrix))

ra <- rowAnnotation(
  Alterations = anno_barplot(
    mut_counts,
    border = FALSE,
    axis_param = list(side = "bottom")
  ),
  annotation_width = unit(1.6, "cm")
)

# Plot
pdf(paste0(path_plots,"OncoPrint_2.pdf"), width = 15, height =20 )
oncoPrint(
  heatmap_matrix,
  alter_fun = alter_fun,
  col = mutation_colors,
  remove_empty_columns = TRUE,
  remove_empty_rows = TRUE,
  show_column_names = TRUE,
  alter_fun_is_vectorized = FALSE,
  show_pct = FALSE,
  show_row_names = FALSE,
  column_title = "Phenotypic Profile of GATA2 Deficiency Cases",
  row_title = "Participants",
  right_annotation = ra
)

 dev.off()
```





```{r}





```






```{r}

library(pheatmap)
heat_mat_child <- heat_mat[wide_df$age_group == "<18", , drop = FALSE]
col_child <- colSums(heat_mat_child == "Y", na.rm = TRUE) > 0
heat_mat_child <- heat_mat_child[ , col_child, drop = FALSE]

heat_mat_adult <- heat_mat[wide_df$age_group == "≥18", , drop = FALSE]
col_adult <- colSums(heat_mat_adult == "Y", na.rm = TRUE) > 0
heat_mat_adult <- heat_mat_adult[ , col_adult, drop = FALSE]


# For pheatmap: map Y/N/NE to 1/2/3
heat_mat_child_num <- matrix(
  dplyr::recode(heat_mat_child, Y = 1L, N = 2L, NE = 3L),
  nrow = nrow(heat_mat_child),
  ncol = ncol(heat_mat_child),
  dimnames = dimnames(heat_mat_child)
)
heat_mat_adult_num <- matrix(
  dplyr::recode(heat_mat_adult, Y = 1L, N = 2L, NE = 3L),
  nrow = nrow(heat_mat_adult),
  ncol = ncol(heat_mat_adult),
  dimnames = dimnames(heat_mat_adult)
)
my_colors <- c("#0072B2", "#D55E00", "#FFFFFF")
names(my_colors) <- c("1", "2", "3")


library(grid)

pdf(paste0(path_plots,"Phenotype_heatmap_Over18.pdf"), width = 15, height =11 )
pheatmap(
  heat_mat_child_num,
  color = my_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  fontsize_col = 8,
  main = paste0("Phenotype presence <18 years (n = ", nrow(heat_mat_child_num), ")"),
  legend = FALSE
)
# After pheatmap, add a custom legend using grid
grid::grid.rect(x = 0.91, y = 0.7, width = 0.12, height = 0.16, gp = gpar(fill = "white", col = NA)) # optional: white background for legend
legend_labels <- c("Y", "N", "NE")
legend_colors <- c("#0072B2", "#D55E00", "#FFFFFF")
y_start <- 0.77
y_gap <- 0.05
for(i in seq_along(legend_labels)) {
  grid::grid.rect(x = 0.85, y = y_start - (i - 1)*y_gap, width = 0.03, height = 0.04,
                  gp = grid::gpar(fill = legend_colors[i], col = "black"))
  grid::grid.text(legend_labels[i], x = 0.89, y = y_start - (i - 1)*y_gap, just = "left", gp = gpar(fontsize = 12))
}
dev.off()


# ADULTS
pdf(paste0(path_plots,"Phenotype_heatmap_Over18.pdf"), width = 15, height =11 )
pheatmap(
  heat_mat_adult_num,
  color = my_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  fontsize_col = 8,
  main = paste0("Phenotype presence ≥18 years (n = ", nrow(heat_mat_adult_num), ")")
)
dev.off()

```
```{r}

pdf(paste0(path_plots,"Phenotype_heatmap_Under18.pdf"), width = 15, height =11 )
pheatmap(
  heat_mat_child_num,
  color = my_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  fontsize_col = 8,
  main = paste0("Phenotype presence <18 years (n = ", nrow(heat_mat_child_num), ")"),
  
)
dev.off()
```

#Convert to Binary
```{r}
# Map Y/N/NE to 1/2/3 for plotting
heat_mat_numeric <- matrix(
  dplyr::recode(heat_mat, Y = 1L, N = -1L, NE = 0L),
  nrow = nrow(heat_mat),
  ncol = ncol(heat_mat),
  dimnames = dimnames(heat_mat)
)


# For pheatmap: map Y/N/NE to 1/0/0
heat_mat_child_num <- matrix(
  dplyr::recode(heat_mat_child, Y = 1L, N = -1L, NE = 0L),
  nrow = nrow(heat_mat_child),
  ncol = ncol(heat_mat_child),
  dimnames = dimnames(heat_mat_child)
)
heat_mat_adult_num <- matrix(
  dplyr::recode(heat_mat_adult, Y = 1L, N = -1L, NE = 0L),
  nrow = nrow(heat_mat_adult),
  ncol = ncol(heat_mat_adult),
  dimnames = dimnames(heat_mat_adult)
)
```

#Keep columns that are both present in UKB and GATA2
```{r}

select_pheno_final<-select_pheno[select_pheno$Present_in_GATA2=="Yes" & select_pheno$Present_in_UKB=="Yes",]


```


#Correlation Matrix for all participants
```{r}
# Assuming your dataframe is called pheno_df
# Columns: phenotype names, Rows: participants, Values: 0/1

# Compute correlation matrix
cor_matrix <- cor(heat_mat_numeric, method = "pearson") # Or use method = "spearman"

# Save as CSV (optional)
write.csv(cor_matrix, paste0(path_plots,"phenotype_correlation_matrix_all_coding2.csv"))

# To get top correlated pairs
library(reshape2)
cor_melt <- melt(cor_matrix)
cor_melt <- cor_melt[cor_melt$Var1 != cor_melt$Var2, ]
cor_melt <- cor_melt[abs(cor_melt$value) > 0.5, ]
cor_melt <- cor_melt[order(-abs(cor_melt$value)), ]
head(cor_melt)


library(ggplot2)

pdf(paste0(path_plots,"Correlation_Plot_119_Labeled_All339_coding2.pdf"),height=10,width=10)
# If cor_melt is already filtered and sorted as above:
ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Top Phenotype-Phenotype Correlations", x = "", y = "")
dev.off()

```


#Correlation Matrix for Adults only

```{r}

# Assuming your dataframe is called pheno_df
# Columns: phenotype names, Rows: participants, Values: 0/1

# Compute correlation matrix
cor_matrix <- cor(heat_mat_adult_num, method = "pearson") # Or use method = "spearman"

# Save as CSV (optional)
write.csv(cor_matrix, paste0(path_plots,"phenotype_correlation_matrix_adult.csv"))

# To get top correlated pairs
library(reshape2)
cor_melt <- melt(cor_matrix)
cor_melt <- cor_melt[cor_melt$Var1 != cor_melt$Var2, ]
cor_melt <- cor_melt[abs(cor_melt$value) > 0.5, ]
cor_melt <- cor_melt[order(-abs(cor_melt$value)), ]
head(cor_melt)


library(ggplot2)

pdf(paste0(path_plots,"Correlation_Plot_119_Labeled_Adult212.pdf"),height=10,width=10)
# If cor_melt is already filtered and sorted as above:
ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Top Phenotype-Phenotype Correlations", x = "", y = "")
dev.off()

```



```{r}

# Assuming your dataframe is called pheno_df
# Columns: phenotype names, Rows: participants, Values: 0/1

# Compute correlation matrix
cor_matrix <- cor(heat_mat_child_num, method = "pearson") # Or use method = "spearman"

# Save as CSV (optional)
write.csv(cor_matrix, paste0(path_plots,"phenotype_correlation_matrix_child.csv"))

# To get top correlated pairs
library(reshape2)
cor_melt <- melt(cor_matrix)
cor_melt <- cor_melt[cor_melt$Var1 != cor_melt$Var2, ]
cor_melt <- cor_melt[abs(cor_melt$value) > 0.5, ]
cor_melt <- cor_melt[order(-abs(cor_melt$value)), ]
head(cor_melt)


library(ggplot2)

pdf(paste0(path_plots,"Correlation_Plot_119_Labeled_Child127.pdf"),height=10,width=10)
# If cor_melt is already filtered and sorted as above:
ggplot(cor_melt,aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Top Phenotype-Phenotype Correlations", x = "", y = "")
dev.off()

```

```{r}
# Ensure rownames(heat_mat) is a column for merging
heat_mat$ParticipantID <- rownames(heat_mat)

# Merge variant type information
heat_mat_merged <- heat_mat %>%
  left_join(
    gata2_survey_data %>% select(prevalence_patient_deidentifier_lab_family_patient, VariantType),
    by = c("ParticipantID" = "prevalence_patient_deidentifier_lab_family_patient")
  )

```






#HEATMAP SPLIT BY ADULT AND PED
```{r}

getwd()
# ─────────────────────────────────────────────────────────────
# 0.  Packages  (install.packages() if needed)
# ─────────────────────────────────────────────────────────────
library(dplyr)
library(tidyr)
library(pheatmap)
library(RColorBrewer)

# ─────────────────────────────────────────────────────────────
# 1.  Mapping table: code  →  description
# ─────────────────────────────────────────────────────────────

mapping_df <- select_pheno %>% dplyr::select(Phenotype, Phenotype_Description)

# ─────────────────────────────────────────────────────────────
# 2.  Identify phenotype columns present in the main data frame
# ─────────────────────────────────────────────────────────────
pheno_cols <- mapping_df$Phenotype         # e.g. "P1"…"P77"

# ─────────────────────────────────────────────────────────────
# 3.  Build a wide binary matrix  (rows = participants)
# ─────────────────────────────────────────────────────────────
wide_df <- phenotype_gata2_og %>%
  # keep only the needed columns
  dplyr::select(Age, all_of(pheno_cols)) %>%
  # turn "Y" into 1, everything else into 0
  mutate(across(all_of(pheno_cols), ~ ifelse(.x == "Y", 1, 0))) %>%
  # add age‑group label
  mutate(age_group = if_else(Age < 18, "<18", "≥18")) 

# ─────────────────────────────────────────────────────────────
# 4.  Order rows: children first, then adults
# ─────────────────────────────────────────────────────────────
wide_df <- wide_df %>% arrange(age_group)

# how many rows go into the first block?   (for the gap)
n_children <- sum(wide_df$age_group == "<18")

# ─────────────────────────────────────────────────────────────
# 5.  Drop helper columns & convert to matrix
# ─────────────────────────────────────────────────────────────
heat_mat <- wide_df %>%
  dplyr::select(all_of(pheno_cols)) %>%
  as.matrix()

# ─────────────────────────────────────────────────────────────
# 6.  Replace column codes by their descriptions
# ─────────────────────────────────────────────────────────────
colnames(heat_mat) <- mapping_df$Phenotype_Description[
                        match(colnames(heat_mat), mapping_df$Phenotype)]

# ─────────────────────────────────────────────────────────────
# 7.  Annotation bar for age
# ─────────────────────────────────────────────────────────────
row_anno <- data.frame(Age = wide_df$age_group)
rownames(row_anno) <- rownames(heat_mat)

# colour palette for 0/1
heat_cols <- colorRampPalette(c("#f7fbff", "#08306b"))(100)

# ─────────────────────────────────────────────────────────────
# 8.  Plot & save to PDF
# ─────────────────────────────────────────────────────────────
pdf("phenotype_heatmap_by_age.pdf", width = 10, height = 8)
pheatmap(heat_mat,
         color            = heat_cols,
         cluster_rows     = TRUE,
         cluster_cols     = TRUE,
         annotation_row   = row_anno,
         annotation_colors = list(Age = c("<18" = "#fc8d59", "≥18" = "#99d594")),
         gaps_row         = n_children,          # horizontal gap between groups
         show_rownames    = FALSE,               # hide long IDs; toggle if desired
         fontsize_col     = 8,
         main             = "Phenotype presence by age group (n = 339)")+coord_flip()
dev.off()

# ─────────────────────────────────────────────────────────────
# 7½.  OPTIONAL FILTER – keep only observed phenotypes
#      (place right after Step 7, just before writing PDFs)
# ─────────────────────────────────────────────────────────────
observed_cols <- colSums(heat_mat, na.rm = TRUE) > 0
heat_mat      <- heat_mat[ , observed_cols, drop = FALSE]

## 8·1  Children (<18)

heat_mat_child <- heat_mat[wide_df$age_group == "<18", , drop = FALSE]
col_child<-colSums(heat_mat_child, na.rm = TRUE) > 0
heat_mat_child<- heat_mat_child[ , col_child, drop = FALSE]

min(colSums(heat_mat_child))
pdf("phenotype_heatmap_under18.pdf", width = 15, height = 8)
r<-pheatmap(heat_mat_child,
         color            = heat_cols,
         cluster_rows     = TRUE,
         cluster_cols     = TRUE,
         show_rownames    = FALSE,
         fontsize_col     = 8,
         main             = paste0("Phenotype presence <18 years (n = ",
                                   nrow(heat_mat_child), ")"))
dev.off()


ggsave("Phenotype_Heatmap_in_Child.png",plot= r ,
       width  = 12, height = 15, units = "in")


## Adult heat‑map
heat_mat_adult <- heat_mat[wide_df$age_group == "≥18", , drop = FALSE]
col_adult<-colSums(heat_mat_adult, na.rm = TRUE) > 0
heat_mat_adult<- heat_mat_adult[ , col_adult, drop = FALSE]


pdf("phenotype_heatmap_adult.pdf", width = 10, height = 8)
pheatmap(heat_mat_adult,
         color        = heat_cols,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = FALSE,
         fontsize_col = 8,
         main = paste0("Phenotype presence ≥18 years (n = ",
                       nrow(heat_mat_adult), ")"))
dev.off()
p<-pheatmap(heat_mat_adult,
         color        = heat_cols,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = FALSE,
         fontsize_col = 8,
         main = paste0("Phenotype presence in Adults (n = ",
                       nrow(heat_mat_adult), ")"))

ggsave("Phenotype_Heatmap_in_Adults.png",plot= p ,
       width  = 12, height = 15, units = "in")

```








